Parallelization of Fast $\ell_1$-Minimization for Face Recognition

### 2011-05-20 main.tex
Some wording fixes in intro

Numerically, the original solution \cite{WagnerA2009-CVPR} was implemented based 
on standard interior-point methods, and its speed has been since improved by ALM in a more recent paper
\cite{WagnerA2011-PAMI}. However, due the high per-class cost
of the alignment step, the recognition system still fails to achieve \emph{real-time recognition} against datasets of hundreds or thousands of users.
In this paper, in addition to accelerating the generic
$\ell_1$-min objectives \eqref{eq:l1min} and \eqref{eq:l1min_denoise}, we 
also discuss how to properly accelerate the iterative face alignment step
\eqref{eq:l1min_alignment} on multi-core CPUs and GPUs.

In this paper we will clearly distinguish between algorithm {\em parallelism},
which is a property of an algorithm, and hardware {\em concurrency}, which is a
property of a hardware architecture. Parallelism in an algorithm provides the
opportunity to perform computations concurrently on hardware.  Algorithms often
exhibit multiple levels of parallelism, and hardware often provides multiple
levels of concurrency.  In these terms, the primary contribution of this paper
will be to determine the optimal mappings between the parallelism available in
$\ell_1$-min and the concurrency available in multi-core CPU and GPU
architectures.

Finally, we present extensive benchmark to compare the performance of the
generic ALM $\ell_1$-min algorithm and the corresponding face recognition
routines on massively parallel CPUs and GPUs.  To this end, a new face
recognition system for security and access control applications has been fully implemented on a state-of-the-art
workstation, as shown in Figure \ref{fig:pipeline}. 
Since the parallelization of any algorithm
is highly dependent on the target hardware, we will give a brief overview of
the architecture of our test system in detail in Section
\ref{sec:concurrency}.
Finally, note that for applications targeting thousands of users or more, it would likely be
necessary to distribute the work over a network of computers, while this paper
focuses on the parallelism available in a single (shared memory) machine.
The C++ source code of our implementations, which target Nvidia's CUDA GPU architecture,
will be made available for peer evaluation.
\begin{figure}
\centering
{\tiny \input{figures/pipeline_simplified.tex}}
\caption{The face recognition pipeline.}
\label{fig:pipeline}
\end{figure}

@@

Although Homotopy can be shown to exactly estimate BP when the solution is
sufficiently sparse \cite{DonohoD2006}, the algorithm still involves 
computationally expensive operations such as matrix-matrix multiplication and 
the solution of least-squares problems with varying $A$ matrices. 
In Section \ref{sec:ALM}, we contend
that ALM is a better choice for implementation on many-core CPUs and GPUs. The ALM algorithm belongs to a
category of approximate $\ell_1$-min solutions called \emph{iterative
shrinkage-thresholding} (IST) methods \cite{WrightS2008,BeckA2009}.
IST algorithms mainly involve elementary operations such as vector 
algebra and matrix-vector multiplication. Therefore,
when the dimension of the problem becomes high, IST-type algorithms are
particularly suitable for hardware systems with a high degree of concurrency. In
\cite{YangA2010-ICIP}, the authors further showed that, compared to other
IST-type algorithms, ALM is able to achieve good estimation accuracy comparable
to BP and Homotopy, while at the same time significantly improve the speed
compared to these exact $\ell_1$-min solutions. Due to these facts, in this
paper, we choose ALM as the core algorithm to implement the complete parallel
face recognition pipeline.

>>>>

Numerically, the original solution \cite{WagnerA2009-CVPR} was implemented based 
on standard interior-point methods, and its speed has been since improved by ALM in a more recent paper
\cite{WagnerA2011-PAMI}. However, due the high per-class cost
of the alignment step, the recognition system still fails to achieve real-time recognition against datasets of hundreds or thousands of users.
In this paper, in addition to accelerating the generic
$\ell_1$-min objectives \eqref{eq:l1min} and \eqref{eq:l1min_denoise}, we 
also discuss how to properly accelerate the iterative face alignment step
\eqref{eq:l1min_alignment} on multi-core CPUs and GPUs.

In this discussion we will clearly distinguish between algorithm {\em parallelism},
which is a property of an algorithm, and hardware {\em concurrency}, which is a
property of a hardware architecture. Parallelism in an algorithm provides the
opportunity to perform computations concurrently on hardware.  Algorithms often
exhibit multiple levels of parallelism, and hardware often provides multiple
levels of concurrency.  In these terms, the primary contribution of this paper
will be to determine the optimal mappings between the parallelism available in
$\ell_1$-min and the concurrency available in multi-core CPU and GPU
architectures.

Finally, we present extensive benchmark to compare the performance of the
generic ALM $\ell_1$-min algorithm and the corresponding face recognition
routines on massively parallel CPUs and GPUs.  To this end, a face
recognition system for security and access control applications has been implemented, 
and is diagrammed in Figure \ref{fig:pipeline}. This pipeline follows the work of \cite{WagnerA2011-PAMI}, but
the fast implementations presented in this paper are novel.
Since the parallelization of any algorithm
is highly dependent on the target hardware, we will give a brief overview of
the architecture of our test system in detail in Section
\ref{sec:concurrency}.
Finally, note that 
while this paper focuses on the parallelism available in a single (shared memory) machine,
applications targeting thousands of users or more, will likely require computation be
distributed over a cluster or network of computers, and this is not addressed here.
The source code of our implementations, which target Nvidia's CUDA GPU architecture,
as well as the x86-64 architecture, will be made available for peer evaluation.
\begin{figure}
\centering
{\tiny \input{figures/pipeline_simplified.tex}}
\caption{The face recognition pipeline.}
\label{fig:pipeline}
\end{figure}

@@

Although Homotopy can be shown to exactly estimate BP when the solution is
sufficiently sparse \cite{DonohoD2006}, the algorithm still involves
computationally expensive operations such as matrix-matrix multiplication and
the solution of least-squares problems with varying $A$
matrices\footnote{During the face alignment algorithm presented later, The $A$
matrix changes between alignment iterations but not within the L1 solver}.  In
Section \ref{sec:ALM}, we contend that ALM is a better choice for
implementation on many-core CPUs and GPUs. The ALM algorithm belongs to a
category of approximate $\ell_1$-min solutions called \emph{iterative
shrinkage-thresholding} (IST) methods \cite{WrightS2008,BeckA2009}.  IST
algorithms mainly involve elementary operations such as vector algebra and
matrix-vector multiplication. Therefore, when the dimension of the problem
becomes high, IST-type algorithms are particularly suitable for hardware
systems with a high degree of concurrency. In \cite{YangA2010-ICIP}, the
authors showed that ALM is able to significantly improve the solver speed,
while still achieving estimation accuracy competitive with exact $\ell_1$-min
solutions, such as BP and Homotopy methods, as well as other IST-type
algorithms.  Therefore, in this paper, we choose ALM as the core algorithm to
implement the complete parallel face recognition pipeline.
